Estimating Real-Time Highstreet Footfall from Wi-Fi Probe Requests

### 2018-11-21 review/response.md

First review comments
=====================

Title : Estimating Real-Time Highstreet Footfall from Wi-Fi Probe Requests

Manuscript id : IJGIS-2018-0340

Comments dated : 28-Oct-2018

@@

1. While this is an important and timing topic dealing with data uncertainty
   and privacy in the mobile age, the reviewers have raised concerns on
	- Clarity of the methodology
	- Lack of discussion of wifi-based human activity studies in the big data
	age
	- Proper use of terminology
	- Performance of the algorithm etc.

2. In lieu of these comments, they have recommended to make substantial
revisions to your manuscript

3. Also, since this is a "smart cities" related special issue, I ask you to
	- Add more discussions on smart cities in the introduction and conclusion
	sections and
	- Link your research more tightly to this theme.

> Thank you for the very helpful comments of the reviewers of our paper/ As
> requested, we have responded to all comments without adding significant
> length to the paper. We have also extended our discussion of how effective
> measurement of footfall is relevant to understanding the functioning of the
> Smart City.
> 
> If any of what follows and what has been uploaded is unclear, please do not
> hesitate to contact me.
> 
> Yours sincerely,
> Balamurugan Soundararaj,
> (on behalf also of Paul Longley and James Cheshire)

@@

8. You need to check that the paper title in ScholarOne is the same as the
title on the paper and that the Title page contains all authors’ names and
affiliations as you would want them to appear.

9. When submitting your revised manuscript, you will be able to respond to the
comments made by the reviewers in the space provided.  You can use this space
to document any changes you make to the original manuscript.  In order to
expedite the processing of the revised manuscript, please be as specific as
possible in your response to the reviewers.

10. Please delete any redundant files before completing the submission.

> Thank you for the detailed instructions. All of these have been checked in
> revised manuscript for the final submission.

@@

2 Reviewer 1
------------

2.1 Overall Summary
-------------------

This paper is about the estimation of pedestrian footfall using Wi-Fi signal
emitted by mobile devices.  It deals with the problem of estimating the
footfall size based on a noisy and biased data. More specifically it addresses
the following issues that may lead to over counting:

- MAC randomization that may cause a device to generate several MAC addresses;
 - Signals received from outside the interest zone; unequal adoption of Wi-Fi
 smartphone in the population.  To address this issue, a number of techniques
 are introduced:
 - clustering received signal strength to remove device that are outside the
 monitored area;
 - clustering probes based on sequence number and timing;
 - finally an adjustment factor is applied to correct the estimated value.

The adjustment factor is calibrated for each location using a manual counting.
The author conducted a pilot experiment to identify the clustering algorithms
and parameters, and then they conducted a second range of experiments at
different location over an extended period of time. During each of those
experiments, ground-truth count was collected manually.  The results show that
the presented approach can significantly reduce the counting error compared to
a naive solution consisting in counting the distinct number of MAC observed.
Nevertheless, there are cases (2) where the error is still large (50%), even if
in other cases (1) the error is only of 9%.

> This is a correct surmisal of the remit of the paper. We have also added
> material to the introduction and conclusion of the paper to better link this
> remit to the Smart Cities discourse.

@@

[1] Mathy Vanhoef, Célestin Matte, Mathieu Cunche, Leonardo Cardoso, Frank
Piessens. Why MAC Address Randomization is not Enough: An Analysis of Wi-Fi
Network Discovery Mechanisms. ACM AsiaCCS, May 2016, Xi’an, China. 2016,

> Thank you. We have included a citiation to the source, as well as one 
> other recent addition to this fast-developing literature.

2. The author considered that the hashing of MAC address is a good way to 
protect users privacy, as some MAC address collected by the device may not be 
random. However hashing of MAC address cannot be considered as a proper 
anonymization technique (see [2] and 
http://webpolicy.org/2014/03/19/questionable-crypto-in-retail-analytics/ ); 
at best it can be considered as an obfuscation. 
[2] Levent Demir, Mathieu Cunche, Cédric Lauradoux. Analysing the privacy 
policies of Wi-Fi trackers. Workshop on Physical Analytics, Jun 2014, Bretton 
Woods, United States. ACM, 2014

> Thank you for pointing this out. In the revised version of the paper
> we have made reference to this discussion, while also making clear the 
> anonymisation is ensured by the system architecture that processess and
> curates the data. Specifically we rotated the random salt used to hash the
> MAC addresses regulary to prevent any kind of tracking of the users across a
> longer period of time with which the hashing process is carried out.

@@

> This is a good point, and is very important for our validation work. We
> have included a more detailed and explicit discussion of the way that field
> observation was used to calibrate and cross validate the Wi-Fi results.

2.3 Clarifications & Corrections
--------------------------------

1. The topic addressed in this work is interesting and timely. The proposed
approach is sound and shows promising results. The fact that the authors
conducted field experiments to evaluate their solution is a positive point.
Yet before publication, I believe that there are a number of points that needs
to be clarified or corrected. Regarding the technical aspects I have a number
of minor remarks:

> Thank you very much for these very positive comments. The paper reports on
> extensive methodological research and field implementation and is, we
> believe, cutting edge in its finding. It also has important implications for
> our understanding of the functioning of smart cities, and these implications
> are discussed in greater detail in the revised version.

@@

8. "The strength of the signal which transmitted the request.". In fact the
Received Signal Strength Indicator, a value providing an indication on the
strength of the received signal as seen by the receiver, not the strength at
the emitter side.

> Thank you. This distinction has been noted in the revised version of the
> paper.

9. Could you provide more information on the application used for counting
(Clicker)

> This simple application was developed specifically for the project for
> Android platform. More details on the application has been included in the
> revised manuscript.

@@

10. "Though the recycling of sequence number r after 4000 leads to multiple
classifications reported on single device, the magnitude of error is greatly
reduced." This sentence is not clear. And the cycle length is not 4000 but 4096
= 2^12.

> Thank you. This point has been boted and accommodated into the revised
> version.

11. Regarding the issue of cycling sequence number, one could use a distance
with a modulo : di,j = di - dj mod 4096.

> Thank you very much for the suggestion. From tests carried out on sample of
> 1910 probe requests from google phones in the pilot study we found that
> around 0.5% reset their mac addresses.  Which in our opinion doesn't
> influence the results in any significant way considering the context.

### 2018-11-21 paper/sections/introduction.tex
Wed 21 Nov 17:08:08 GMT 2018

Sentient technologies include mobile phone networks, which can triangulate user
locations relative to networks of masts, use of GPS to locate users of social
media services, and Wi-Fi beacon connectivity to access the Internet. These
technologies offer differing levels of spatial precision, where mobile telephony
and Wi-Fi generally being less reliable and offer lower precision than GPS to
the end users, while being more advantageous for broader mobility studies
(Pinelli et al.  2015). There has been considerable research into the usefulness
of these technologies to understand patterns of movement in cities in near real
time (Candia et al. 2008, Gonzalez et al. 2008, Calabrese et al. 2013).  Most of
this research has focused upon technical specification of accuracy or precision
(Song et al. 2010, Lane et al. 2010), with somewhat less attention devoted to
the ways that the characteristics of the technologies and of their human users
conspire to create possible bias in representing usage patterns across the
entire smart city.

>>>>

Sentient technologies include mobile phone networks, which can triangulate user
locations relative to networks of masts, use of GPS to locate users of social
media services, and Wi-Fi access points connectivity to access the Internet.
These technologies offer differing levels of spatial precision, where mobile
telephony and Wi-Fi generally being less reliable and offer lower precision than
GPS to the end users, while being more advantageous for broader mobility studies
(Pinelli et al.  2015). There has been considerable research into the usefulness
of these technologies to understand patterns of movement in cities in near real
time (Candia et al. 2008, Gonzalez et al. 2008, Calabrese et al. 2013).  Most of
this research has focused upon technical specification of accuracy or precision
(Song et al. 2010, Lane et al. 2010), with somewhat less attention devoted to
the ways that the characteristics of the technologies and of their human users
conspire to create possible bias in representing usage patterns across the
entire smart city.

### 2018-11-21 paper/sections/methodology.tex
Wed 21 Nov 17:08:08 GMT 2018

Probe requests are a special type of management packets broadcast by Wi-Fi
enabled devices as part of the various functions such as scanning for available
APs, quick geolocation by triangulation based known APs, etc. These are
broadcast by all Wi-Fi enabled devices regardless of the manufacturer, type or
model of the devices though there is some variation on the frequency and the
information transmitted through them. In some cases, such as Android devices,
these are broadcast even when the Wi-Fi functionality has been turned off by the
user. Thus these signals can be used to reliably identify the presence of Wi-Fi
enabled mobile devices. Being a first step of connection initiated by the mobile
device, these packets have information regarding the characteristics of the
mobile device itself. Some of the key information we can infer from these
requests are,

\begin{enumerate} 
\item \textbf{Media Access Control (MAC) address} which is an unique identifier
	for the wireless hardware of the mobile device, 
\item \textbf{Sequence number} of the request for the mobile device to keep
	track of the responses, 
\item \textbf{Time stamp} at which the request was received by the AP, 
\item Total \textbf{length} of the request in number of bits, and 
\item The \textbf{strength of the signal} which transmitted the request.
\end{enumerate}

The MAC address is the primary unique identifier for the mobile device. It has
two parts, first part is the Organisation Unique Identifier (OUI) which gives
information about the manufacturer of the device and the second is unique to the
device. In modern devices, to protect the users privacy, the MAC address can
also be randomised (hence non unique) and is marked as such. Though sequence
number and length of the packet are not strictly unique, we hypothesize that we
can use them to estimate the number unique devices as demonstrated by
\citep{vanhoef2016}.

Data collection was done with the help of custom sensors built from modifying
the hardware used in Smart street sensors \citep{sss2016} and updating them with
custom software. The sensor is essentially a Raspberry-Pi device with Wi-Fi and
3G modules. It keeps the Wi-Fi module in `Monitor' mode and uses the open source
software - Wireshark \citep{wireshark2} to passively collect all packets sent to
`broadcast', marked with type as management' and subtype as `probe requests'.
The MAC address in these probe requests is anonymised at the device level using
a cryptographic hashing algorithm and transmitted through 3G connection to a
central database via web-sockets protocol, where it is stored in a PostgreSQL
database for further analysis. A overall schematic of the data collection and
storage process is shown in Figure \ref{datacollection_schematic}. The ground
truth on number of pedestrian footfall was recorded using a custom mobile
application - Clicker \citep{bala2018clicker}

@@

We hypothesise that in configurations where a specific source of background
noise is at a constant distance, there must be a distinct pattern in the number
of probe requests reporting signal strength corresponding to that distance. For
example, if there is a phone shop next to our sensor where hundreds of phones
regularly send probe requests there should be a sharp rise of number of probe
requests with reported signal strength corresponding to the distance between the
sensor and the phone shop irrespective of the local conditions as shown in
Figure \ref{methodology_schematic}. We could identify these breaks in the data
using traditional one dimensional clustering algorithms such as `jenks natural
breaks', `k-means', `quantile' and `hierarchical clustering', etc. Since we are
only looking for the break in the data and not for absolute values, the
methodology should apply for all the variations due micro site conditions thus
reducing the overall noise in the collected data.

@@

From our initial analysis we found that OUI, length of the packet and sequence
number of the packet being the most promising information to achieve this.
First we divide our dataset into sets of probe requests with randomised and
non-randomised MAC addresses and keep the MAC address as the unique identifier
for the latter set. For randomised ones we further divide them in to sub
categories based on their OUI and length of the packet. Since the length tends
to stay unique to specific models of devices we are left with the task of
identifying the unique mobile devices from within these sets.

@@

Since mobile device ownership is an external uncertainty to our study and could
arise from variety of spatio - temporal and demographic factors, we propose to
solve this using external source of information. We hypothesize that an
adjustment factor could be arrived at for each location of data collection by
comparing the sensor based counts and ground truth and it can be used to adjust
the data reliably to reflect the ground truth in absolute numbers for the
future. This calibration can be carried out periodically and frequently to
improve the quality of the estimation over longer periods of time.

>>>>

Probe requests are a special type of management packets broadcast by Wi-Fi
enabled devices as part of the various functions such as scanning for available
APs, quick geolocation by triangulation based known APs, etc. These are
broadcast by all Wi-Fi enabled devices regardless of the manufacturer, type or
model of the devices though there is some variation on the frequency and the
information transmitted through them. In some cases, such as Android devices,
these are broadcast even when the Wi-Fi functionality has been turned off by
the user so that device can immediately connet to networks when the
functionality is switched back on. Since some devices even use the probe
requests as a less accurate form of localisation they keep sending probe
requests as well. Thus these signals can be used to reliably identify the
presence of Wi-Fi enabled mobile devices. Being a first step of connection
initiated by the mobile device, these packets have information regarding the
characteristics of the mobile device itself. Some of the key information we can
infer from these requests are,

\begin{enumerate} 
\item \textbf{Media Access Control (MAC) address} which is an name identifying
	for the wireless hardware of the mobile device, 
\item \textbf{Sequence number} of the request for the mobile device to keep
	track of the responses, 
\item \textbf{Time stamp} at which the request was received by the AP, 
\item Total \textbf{length} of the request in number of bits, and 
\item The \textbf{strength of the signal} received by the mobile device.
\end{enumerate}

The MAC address is the primary unique identifier for the mobile device. It has
two parts, first part is the Organisationally Unique Identifier (OUI) which
gives information about the manufacturer of the device and the second is unique
to the device. In modern devices, to protect the users privacy, the MAC address
can also be randomised (hence non unique) and is marked as such. Though
sequence number of the packet is strictly unique to a mobile device, we
hypothesize that we can use them to estimate the number unique devices as
demonstrated by \citep{vanhoef2016}, where optional information present in the
probe requests - Information Elements (IE) along with the probe requests have
been used to fingerprint the devices. This approach has become increasinly
difficult as mobile phone manufacturers have severely limited the IEs present
in the probe requests thus leading us to explore methods which uses only the
sequence numbers. This also affects the established commercial solutions using
Wi-Fi probe requests such as Blix, Walkbase, Euclid Analytics, RetailNext etc.
There have been another solution proposed by \citep{hong2018crowdprobe} where
the authors tried to solve the similar problem using a hidden markov models
based trajectory inference alogorithm but the scope is limited to enclosed,
exit controlled public spaces such as shopping malls, railway stations, etc.

Data collection was done with the help of custom sensors built from modifying
the hardware used in Smart street sensors \citep{sss2016} and updating them with
custom software. The sensor is essentially a Raspberry-Pi device with Wi-Fi and
3G modules. It keeps the Wi-Fi module in `Monitor' mode and uses the open source
software - Wireshark \citep{wireshark2} to passively collect all packets sent to
`broadcast', marked with type as management' and subtype as `probe requests'.
The MAC address in these probe requests is obfuscated at the device level using
a cryptographic hashing algorithm and transmitted through 3G connection to a
central database via web-sockets protocol, where it is stored in a PostgreSQL
database for further analysis. The random salt used in the hashing algorithm was
rotated regularly to further mitigate the risk of de-anonymisation of the hash.
Though hashing cannot completely ensure anonymisation as discussed in
\citep{demir2014analysing} it can sufficiently obfuscate the data which along
with a secure process of data handling gives us reasonable security. A overall
schematic of the data collection and storage process is shown in Figure
\ref{datacollection_schematic}. The ground truth on number of pedestrian
footfall was recorded using a custom Android application - Clicker
\citep{bala2018clicker}. This app logs accurate timestamps for each time the
surveryor records a pedestrian crossing the designated cordon line at the
location. In addition to counting the pedestrians manually this also makes the
mobile device used constantly probe for networks which also gives us a know
device to calibrate our methodology against.

@@

We assume that in configurations where a specific source of background
noise is at a constant distance, there must be a distinct pattern in the number
of probe requests reporting signal strength corresponding to that distance. For
example, if there is a phone shop next to our sensor where hundreds of phones
regularly send probe requests there should be a sharp rise of number of probe
requests with reported signal strength corresponding to the distance between the
sensor and the phone shop irrespective of the local conditions as shown in
Figure \ref{methodology_schematic}. We could identify these breaks in the data
using traditional one dimensional clustering algorithms such as `jenks natural
breaks', `k-means', `quantile' and `hierarchical clustering', etc. Since we are
only looking for the break in the data and not for absolute values, the
methodology should apply for all the variations due micro site conditions thus
reducing the overall noise in the collected data.

@@

From our initial look at the data we found that OUI and the sequence
number of the packet is the most promising information to achieve this.
First we divide our dataset into sets of probe requests with randomised and
non-randomised MAC addresses by looking at the second character of the vendor
part of the MAC address and if it is E, A, 2 or 6, then these addresses are
identified to be randomised. We keep the MAC address as the unique
identifier for the non-randomised requests and further divide the randomised ones in to
sub categories based on their OUI. We then  
identify unique mobile devices from within these sets and assign a unique identifier to each device.

@@

Since mobile device ownership is an external uncertainty to our study and could
arise from variety of spatio - temporal and demographic factors, we propose to
solve this using a sample count done manually at each location.  We can then
calculate an  adjustment factor or an offset for each location by comparing the
sensor based counts and ground truth and in turn it can be used to adjust the
data reliably to reflect the ground truth in absolute numbers for the future.
This calibration can be carried out periodically at these locations to improve
the quality of the estimation.

### 2018-11-21 paper/sections/pilot_study.tex
Wed 21 Nov 17:08:08 GMT 2018

\begin{table}
\tbl{Comparison of clustering algorithms with a sample of 40000 probe requests}
{\begin{tabular}{lcc} 
\toprule
	 Algorithm					& Time (s) 	& MAPE (\%) \\
\midrule
	 Quantile					& 0.002 	&  27\% \\
	 K-Means			 		& 0.007 	& -23\% \\
	 Hierarchical Clustering	& 172.520 	& -9\% \\
	 Bagged Clustering 			& 0.135 	& -30\% \\
	 Fisher 					& 3.034 	& -30\% \\
	 Jenks Natural Break 		& 556.279 	& -30\% \\
 \bottomrule
\end{tabular}}
\label{classification-table}
\end{table}

To start we designed a small pilot study to validate the filtering and
clustering methodology against the scale and complexity of data collected on an
open public area such as a retail high street.  We also aim to find the
algorithm which is best suited for classification of signal strengths as 'low'
and 'high' to filter out the background noise.  The data was collected at Oxford
Street, London on 20 December 2017 from 12:30 to 13:00 hrs, where Wi-Fi probe
requests were collected using the sensor described in Chapter \ref{methodology}
and pedestrian footfall was manually recorded using the Android app
\citep{bala2018clicker}.  Being located at one of the busiest retail locations
in the United Kingdom, the Wi-Fi sensor captured approximately 60,000 probe
requests during the half hour period, and 3,722 people were recorded manually
walking on the sidewalk during that time.

>>>>

\begin{table}
\tbl{Comparison of clustering algorithms with a sample of 40000 probe requests}
{\begin{tabular}{lcc} 
	\toprule
	 Algorithm					& Time (s) 	& MAPE (\%) \\
	\midrule
	 Quantile					& 0.002 	&  27 \% \\
	 K-Means			 		& 0.007 	& -23 \% \\
	 Hierarchical Clustering	& 172.520 	&  -9 \% \\
	 Bagged Clustering 			& 0.135 	& -30 \% \\
	 Fisher 					& 3.034 	& -30 \% \\
	 Jenks Natural Break 		& 556.279 	& -30 \% \\
	 \bottomrule
\end{tabular}}
\label{classification-table}
\end{table}

To start we designed a small pilot study to validate the filtering and
clustering methodology against the scale and complexity of data collected on an
open public area such as a retail high street.  We also aim to find the
algorithm which is best suited for classification of signal strengths as 'low'
and 'high' to filter out the background noise.  The data was collected at Oxford
Street, London on 20 December 2017 from 12:30 to 13:00 hrs, where Wi-Fi probe
requests were collected using the sensor described in Chapter \ref{methodology}
and pedestrian footfall was manually recorded using the Android app
\citep{bala2018clicker}.  Being located at one of the busiest retail locations
in the United Kingdom, the Wi-Fi sensor captured approximately 60,000 probe
requests during the half hour period, and 3,722 people were recorded manually
walking on the sidewalk during that time. The count was done at the location in
front of the store where the surveyor carried the sensor on a backpack for the
automated count and used a phone for doing the manual count. The sensor is kept
as close to the store window as possible and the manual count is done as a
cordon count in fron of the store.

### 2018-11-21 paper/sections/previous_work.tex
Wed 21 Nov 17:08:08 GMT 2018

In the past decade Wi-Fi has emerged as one of the most commonly used technology
in providing high speed internet access to mobile devices such as smartphones,
tablets and laptops in public and private spaces. This has resulted in multiple
Wi-Fi networks being available at almost every location in dense urban
environments. Traversing through this overlapping mesh of Wi-Fi networks, modern
mobile devices with Wi-Fi antennae regularly broadcast a special type of signal
known as `Probe Requests' in order to discover Wi-Fi networks available to them.
This helps these devices to connect and switch between the Wi-Fi networks
seamlessly.

Probe requests are low level signals standardised by IEEE 802.1b/g specification
\citep{ieee2013} as the first step in establishing a Wi-Fi based connection
between two devices and is implemented in any Wi-Fi capable device irrespective
of the manufacturer or the model. This ubiquity and standardisation makes them
an excellent source of open, passive, continuous, and wireless data generated by
Wi-Fi capable devices present at any given time and location. Considering the
unprecendented levels of mobile device ownership in recent years, we can, in
turn use this data to understand the population distribution in highly dynamic
urban environments with high spatial and temporal granularity
\citep{freud2015,konto2017}. While a Wi-Fi based method to collect data offers
us various advantages such as, easy scalability and efficiency in terms of cost
and time, It also introduces few systematic biases, uncertainities in the
collected data along with the serious risk of infringing on the privacy of the
mobile users. In this paper, using a set of probe requests and manual counts
collected at various high street locations across London, we demonstrate that
pedestrian footfall at these locations can be estimated with considerable
precision and accuracy while protecting the privacy of the pedestrians.

>>>>

In the past decade Wi-Fi has emerged as one of the most commonly used technology
in providing high speed internet access to mobile devices such as smartphones,
tablets and laptops in public and private spaces. This has resulted in multiple
Wi-Fi networks being available at almost every location in dense urban
environments. Traversing through this overlapping mesh of Wi-Fi networks, modern
mobile devices with Wi-Fi network interface regularly broadcast a special type
of signal known as `Probe Requests' in order to discover Wi-Fi networks
available to them.  This helps these devices to connect and switch between the
Wi-Fi networks seamlessly.

Probe requests are low level signals standardised by IEEE 802.11 specification
\citep{ieee2016} for service discovery and is implemented in any Wi-Fi capable
device irrespective of the manufacturer or the model. This ubiquity and
standardisation makes them an excellent source of open, passive, continuous,
and wireless data generated by Wi-Fi capable devices present at any given time
and location. Considering the unprecendented levels of mobile device ownership
in recent years, we can, in turn use this data to understand the population
distribution in highly dynamic urban environments with high spatial and
temporal granularity \citep{freud2015,konto2017}. While a Wi-Fi based method to
collect data offers us various advantages such as, easy scalability and
efficiency in terms of cost and time, It also introduces few systematic biases,
uncertainities in the collected data along with the serious risk of infringing
on the privacy of the mobile users. In this paper, using a set of probe
requests and manual counts collected at various high street locations across
London, we demonstrate that pedestrian footfall at these locations can be
estimated with considerable precision and accuracy while protecting the privacy
of the pedestrians.
